Estimating duration in the presence of misclassification

### 2024-03-29 biology-data.tex
proof

SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV and MERS-CoV.
Coronaviruses are RNA (Ribonucleic acid) viruses.
RNA viruses use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.

@@

The PCR process is a series of cycles, each cycle containing three steps~\autocite{powledgePCR,garibyanPCR}.
The first step, \emph{denaturation}, heats the DNA causing its two strands to separate.
The second step, \emph{annealing}, lowers the temperature, causing the primers to bind to one of the strands of DNA.
The final step, \emph{extension}, raises the temperature, causing the DNA polymerase to replicate the DNA sequence between the primers. 
In the course of such a cycle, the amount of DNA doubles.
Therefore, the number of DNA copies increases geometrically with the number of cycles completed.

@@

Like any diagnostic test, PCR tests are not perfect.
The quantitative result (the Ct value) has random variation even if the sample has the same initial amount of DNA or RNA present, and the qualitative result (positive or negative) can be wrong.
Errors arise from a variety of sources, including the sampling/swabbing process and batch effects in the RT-PCR process~\autocite{hanRTPCR}.

@@

False positives are very rare for SARS-CoV-2 RT-PCR tests.
In summer 2020, the CIS performed\numprint{208730} PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
If these were all false positives, then the test specificity would be 99.9\%.
In reality, it is almost certain that some of these are true positives, and hence this is a lower-bound on the true test specificity.

@@

The CIS~\autocite{CIS} was set up in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
It was a longitudinal study with a representative sample of private households, \ie excluding settings such as care homes and student halls.
The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.
Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
Enrolment was continuous until 31 Jan 2022, with data collected until 13 Mar 2023~\autocite{weiRisk}.

@@

Up until July 2020, households were selected from those that had previously participated in an ONS survey~\autocite{CIStechData}.
Around 50\% of households agreed to participate, with around 90\% of eligible individuals (\numprint{96113} in total) in these households contributing multiple swabs to the survey.
After July 2020, addresses were also randomly sampled from a database of all UK addresses held by the ONS.
The response rate was lower, with only 12\% of households agreeing to participate, from which 86\% of eligible individuals (\numprint{383732} in total) contributed multiple swabs to the survey.

@@

If the CIS was truly a random sample, an unbiased estimator of the positivity rate would be the proportion of positive tests in the sample; I refer to this estimator as the unadjusted positivity rate.
However, in practice, two issues arise with this.
First, the non-response (discussed above) causes bias.
Second, the number of positives on any day is low, especially among any specific subpopulation (\eg age group or region) increases variance.
Statistical methods are used to correct for these issues.

@@

I modified the regression component of a pre-existing version of a MRP model for the CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for the specific context of this thesis~\citePersonalComms{Koen Pouwels}.
As in prior work~\autocite{pouwelsMRPvaccination}, I used INLA for inference, implemented in R-INLA (see \cref{E-transmission:sec:INLA}).
For a day $t$, age group $a$, ethnicity $e$, and sex $s$ the regression model is:
\begin{align}
    \logit(p_{taes}) &= \beta + \beta_s + \beta_{w(e)} + u_a + v_e + w_t + l_{a,t} \\
    y_{taes} &\dist \BB(n_{taes}, p_{taes}, \rho).
\end{align}
The terms are as follows.
\todo[inline]{Check I've written the priors correctly}
\begin{itemize}
    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results respectively on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$. I use a beta-binomial distribution for the likelihood, as discussed in \cref{biology-data:sec:clustering}.
    \item $\beta$ is a global intercept term with the vague prior $\beta \dist \N(0, 1/0.001^2)$.
    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females. The prior for $\beta_s$ is vague, $\beta_s \dist \N(0, 1/0.001^2)$.
    \item $\beta_{w(e)}$ is a fixed effect for White or non-White ethnicity. $\beta_{w(e)} = 0$ if $e$ corresponds to White ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-White ethnic groups).
      The prior for $\beta_n$ is vague, $\beta_n \dist \N(0, 1/0.001^2)$.
      % $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
    \item $u_a$ and $v_e$ are independent normal random effects for age and ethnic group respectively. Their standard deviation has a penalized complexity prior~\autocite{simpsonPenalising} with a 10\% probability of being greater than 1.
    \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
      $\sigma_w$ has a vague prior, $1/\sigma_w \dist \GamDist(1, 5 \times 10^{-5})$.
    \item $l_{a,t}$ is an age-group specific second-order random walk.
      That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{a,t} \dist \N(0, \sigma_l^2)$.
      $\sigma_l$ has a vague prior, $1/\sigma_l \dist \GamDist(1, 5 \times 10^{-5})$.
\end{itemize}
% For parameters given vague priors, the large amount of data means that the prior has little influence on the posterior distribution.
The second-order random walk priors favours linear growth in $\logit(p_{taes})$ over time.
For the small values of $p_{taes}$ likely to be encountered, this approximates exponential growth.

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
Denote these by $p^(j)_{t,i}$ for the $j$th sample where $i$ is a stratum, \ie some value of $a$, $e$, and $s$.
Define a (sub) population of interest as a set of strata $\set{S}$.
Then the prevalence in the (sub)population on day $t$ is
\begin{align}
p^{(j)}_t(\set{S})
&= \frac{\sum_{i \in \set{S}} p_{t,i} N_j}{\sum_{i \in \set{S}} N_i}
\end{align}
where $N_j$ is the population size of stratum $j$.
The population numbers for each stratum are provided by the Office for National Statistics.

@@

A negative between two positive tests, when those positives are part of the same infection episode, is known as an \emph{intermittent negative}.
Intermittent negatives are the clearest example of a false negative.
I will generally consider that intermittent negatives are uninformative on the duration of detectability, and hence exclude them from several analyses.

The crucial features of an infection episode are the start and end times of the infection.
Ignoring misclassified test results, latent start time of infection episode $j$, $B_j$, can be bounded as between the day after the last negative test, $l_j^{(b)}$, and the day of the first positive test, $r_j^{(b)}$.
Similarly, the latent end time of infection episode $i$, $E_j$, can be bounded as between the day of the last positive test, $l_j^{(e)}$, and the day before the following negative test, $r_j^{(e)}$.

@@

For privacy reasons, the CIS data are stored in the ONS's SRS (Secure Research Service)~\autocite{onsSRS}.
The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.
The SRS gives accredited researchers working on projects in the public interest access to granular data which cannot be legally shared publicly.

>>>>

SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV and MERS-CoV.
Coronaviruses are RNA (Ribonucleic acid) viruses which means they use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.

@@

The PCR process is a series of cycles, each cycle containing three steps~\autocite{powledgePCR,garibyanPCR}.
The first step, \emph{denaturation}, heats the DNA causing its two strands to separate.
The second step, \emph{annealing}, lowers the temperature, causing the primers to bind to one of the strands of DNA.
The final step, \emph{extension}, raises the temperature, causing the DNA polymerase to replicate the DNA sequence between the primers. 
In the course of such a cycle, the amount of target DNA doubles.
Therefore, the number of DNA copies increases geometrically with the number of cycles completed.

@@

Like any diagnostic test, PCR tests are not perfect.
The quantitative result (the Ct value) has random variation even if the sample has the same initial amount of DNA or RNA present, and the qualitative result (positive or negative) can be wrong.
Random variation arises from a variety of sources, including the sampling/swabbing process and batch effects in the RT-PCR process~\autocite{hanRTPCR}.

@@

False positives are very rare for SARS-CoV-2 RT-PCR tests.
In summer 2020, the CIS performed \numprint{208730} PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
If these were all false positives, then the test specificity would be 99.9\%.
In reality, it is almost certain that some of these are true positives, and hence this is a lower-bound on the true test specificity.

@@

The CIS~\autocite{CIS} was set up in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
It was a longitudinal study with a representative sample of private households, \ie excluding settings such as care homes and student halls.
Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
Enrolment was continuous until 31 Jan 2022, with data collected until 13 Mar 2023~\autocite{weiRisk}. 
The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.

@@

Up until July 2020, households were selected from those that had previously participated in an ONS survey~\autocite{CIStechData}.
In this initial phase, around 50\% of households agreed to participate, with around 90\% of eligible individuals (\numprint{96113} in total) in these households contributing multiple swabs to the survey.
After July 2020, addresses were also randomly sampled from a database of all UK addresses held by the ONS.
In this phase, response rate was lower, with only 12\% of households agreeing to participate, from which 86\% of eligible individuals (\numprint{383732} in total) contributed multiple swabs to the survey.

@@

If the CIS was truly a random sample, an unbiased estimator of the population positivity rate would be the proportion of positive tests in the sample; I refer to this estimator as the unadjusted positivity rate.
However, in practice, two issues arise with this.
First, the non-response (discussed above) causes bias.
Second, the number of positives on any day is low, especially among any specific subpopulation (\eg age group or region) increases variance.
Statistical methods are used to correct for these issues.

@@

I modified the regression component of a pre-existing version of a MRP model for the CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for the specific context of this thesis~\citePersonalComms{Koen Pouwels}.
For a day $t$, age group $a$, ethnicity $e$, and sex $s$ the regression model is:
\begin{align}
    \logit(p_{taes}) &= \beta + \beta_s + \beta_{w(e)} + u_a + v_e + w_t + l_{a,t} \\
    y_{taes} &\dist \BB(n_{taes}, p_{taes}, \rho).
\end{align}
The terms are as follows.
\todo[inline]{Check I've written the priors correctly}
\begin{itemize}
    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results respectively on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$. I use a beta-binomial distribution for the likelihood, as discussed in \cref{biology-data:sec:clustering}.
    \item $\beta$ is a global intercept term with the vague prior $\beta \dist \N(0, 1/0.001^2)$.
    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females. The prior for $\beta_s$ is vague, $\beta_s \dist \N(0, 1/0.001^2)$.
    \item $\beta_{w(e)}$ is a fixed effect for White or non-White ethnicity. $\beta_{w(e)} = 0$ if $e$ corresponds to White ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-White ethnic groups).
      The prior for $\beta_n$ is vague, $\beta_n \dist \N(0, 1/0.001^2)$.
      % $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
    \item $u_a$ and $v_e$ are independent normal random effects for age and ethnic group respectively. Their standard deviation has a penalized complexity prior~\autocite{simpsonPenalising} with a 10\% probability of being greater than 1.
    \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
      $\sigma_w$ has a vague prior, $1/\sigma_w \dist \GamDist(1, 5 \times 10^{-5})$.
    \item $l_{a,t}$ is an age-group specific second-order random walk.
      That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{a,t} \dist \N(0, \sigma_l^2)$.
      $\sigma_l$ has a vague prior, $1/\sigma_l \dist \GamDist(1, 5 \times 10^{-5})$.
\end{itemize}
% For parameters given vague priors, the large amount of data means that the prior has little influence on the posterior distribution.
The second-order random walk priors favours linear growth in $\logit(p_{taes})$ over time.
For the small values of $p_{taes}$ likely to be encountered, this approximates exponential growth.

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
Denote these by $p^{(j)}_{t,i}$ for the $j$th sample where $i$ is a stratum, \ie some value of $a$, $e$, and $s$.
Define a (sub) population of interest as a set of strata $\set{S}$.
Then the prevalence in the (sub)population on day $t$ is
\begin{align}
p^{(j)}_t(\set{S})
&= \frac{\sum_{i \in \set{S}} p^{(j)}_{t,i} N_i}{\sum_{i \in \set{S}} N_i}
\end{align}
where $N_i$ is the population size of stratum $i$.
The population numbers for each stratum are provided by the ONS.

@@

A negative between two positive tests, when those positives are part of the same infection episode, is known as an \emph{intermittent negative}.
Intermittent negatives are the clearest example of a false negative.

The crucial features of an infection episode are the start and end times of the infection.
Ignoring misclassified test results, the latent start time of infection episode $j$, $B_j$, can be bounded as between the day after the last negative test, $l_j^{(b)}$, and the day of the first positive test, $r_j^{(b)}$.
Similarly, the latent end time of infection episode $j$, $E_j$, can be bounded as between the day of the last positive test, $l_j^{(e)}$, and the day before the following negative test, $r_j^{(e)}$.

@@

For privacy reasons, the CIS data are stored in the ONS's SRS (Secure Research Service)~\autocite{onsSRS}.
The SRS is a TRE (Trusted Research Environment) that allows researchers to access sensitive data in a secure environment.
The SRS gives accredited researchers working on projects in the public interest access to granular data which cannot be legally shared publicly.
